---
title: "TB Vaccination"
output: 
  - pdf_document
  - html_notebook
---

PDF Version available [here](http://rwhite.no/tb_vaccination/tb_vaccination.pdf).

## Data creation

```{r, message=FALSE, warning=FALSE}
library(data.table)
library(pomp)
library(ggplot2)
library(rstan)

rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

#everyone enters on jan 1
#results are for dec 31
newPeople <- data.table(id=1:1000,year=2000,age=1,timeNoVax=0,timeWithVax=0,vaccinated=0,ltbi=0,atbi=0,ageAtTB=0)

pomp::bake("data/peopleNoVax.RDS",{
  set.seed(4)
  peopleNoVax <- vector("list",length=17)
  peopleNoVax[[1]] <- newPeople[1,][-1,]
  for(i in 1:17){
    # NO VACCINATIONS
    if(i!=1) peopleNoVax[[i]] <- copy(peopleNoVax[[i-1]])
    peopleNoVax[[i]][,age:=age+1]
    peopleNoVax[[i]] <- rbind(peopleNoVax[[i]],newPeople)
    peopleNoVax[[i]][,year:=2000+i]
    
    N <- nrow(peopleNoVax[[i]])
    peopleNoVax[[i]][sample(1:N,size=round(N*0.050)),ltbi:=1]
    peopleNoVax[[i]][,timeNoVax:=timeNoVax+1]
    
    Nltbi <- sum(peopleNoVax[[i]]$ltbi)
    peopleNoVax[[i]][atbi==1,ltbi:=0]
    peopleNoVax[[i]][atbi==1,ageAtTB:=age]
    peopleNoVax[[i]][,atbi:=0] # people with ATBI go to the doctor and get cured last year
    peopleNoVax[[i]][sample(which(peopleNoVax[[i]]$ltbi==1),size=round(Nltbi*0.1)),atbi:=1]
  }
  peopleNoVax
}) -> peopleNoVax

pomp::bake("data/peopleWithVax.RDS",{
  set.seed(4)
  peopleWithVax <- vector("list",length=17)
  peopleWithVax[[1]] <- newPeople[1,][-1,]
  for(i in 1:17){
    # WITH VACCINATIONS, BUT STOPPING THEM AFTERWARDS
    if(i!=1) peopleWithVax[[i]] <- copy(peopleWithVax[[i-1]])
    peopleWithVax[[i]][,age:=age+1]
    peopleWithVax[[i]] <- rbind(peopleWithVax[[i]],newPeople)
    peopleWithVax[[i]][,year:=2000+i]
    N <- nrow(peopleWithVax[[i]])
    if(i>=10){
      peopleWithVax[[i]][sample(1:N,size=round(N*0.050)),ltbi:=1]
      peopleWithVax[[i]][,timeNoVax:=timeNoVax+1]
    } else {
      peopleWithVax[[i]][sample(1:N,size=round(N*0.050)),ltbi:=1]
      peopleWithVax[[i]][,timeWithVax:=timeWithVax+1]
    }
    
    Nltbi <- sum(peopleWithVax[[i]]$ltbi)
    peopleWithVax[[i]] <- peopleWithVax[[i]][atbi==0]
    peopleWithVax[[i]][atbi==1,ltbi:=0]
    peopleWithVax[[i]][atbi==1,ageAtTB:=age]
    peopleWithVax[[i]][,atbi:=0] # people with ATBI go to the doctor and get cured last year
    peopleWithVax[[i]][sample(which(peopleWithVax[[i]]$ltbi==1),size=round(Nltbi*0.1)),atbi:=1]
  }
  peopleWithVax
}) -> peopleWithVax
```

## Detecting LTBI without vaccination

Lets try an easy model, detecting LTBI in unvaccinated people.

We have a Poisson binomial distribution, where the probability of person $i$ not having LTBI is:

$$Pr(Y_i=0 | p, T_i) = (1-p)^{T_i}$$

And the probability of person $i$ having LTBI is:

$$Pr(Y_i=1 | p, T_i) = 1 - Pr(Y_i=0 | p, T_i)$$

```{r}
stanData <- copy(peopleNoVax[[17]])
stanData[ageAtTB>0,ltbi:=1]
stanData <- stanData[,.(R=.N),by=.(timeNoVax,timeWithVax,ltbi)]
data = list(N=nrow(stanData),
            R=stanData$R,
            y=stanData$ltbi*stanData$R,
            timeNoVax=stanData$timeNoVax)

stan_code = "
data {
  int<lower=0> N;
  int R[N];
  int y[N];
  vector[N] timeNoVax;
}
parameters {
  real<lower=0,upper=0.5> thetaNoVax;
}
model {
  thetaNoVax ~ beta(0.05, 1);
  
  for (n in 1:N){
    y[n] ~ binomial(R[n],1-((1-thetaNoVax)^timeNoVax[n]));
  }
}
"
pomp::bake("results/noVaxLTBI.RDS",{
  stan(model_code=stan_code,
              model_name="noVaxLTBI",
              data=data,
              iter=10000, chains=4, init=0, seed=4)
}) -> fit
summary(fit)$summary
```

## Detecting LTBI with vaccination

Lets try a harder model, detecting LTBI when vaccination originally exists, and is then phased out.

We have a Poisson binomial distribution, where the probability of person not $i$ having LTBI is:

$$Pr(Y_i=0 | p_w, p_n, TW_i, TN_i) = (1-p_w)^{TW_i} \times (1-p_n)^{TN_i}$$

And the probability of person $i$ having LTBI is:

$$Pr(Y_i=1 | p_w, p_n, TW_i, TN_i) = 1 - Pr(Y_i=0 | p_w, p_n, TW_i, TN_i)$$

```{r}
stanData <- copy(peopleWithVax[[17]])
stanData[ageAtTB>0,ltbi:=1]
stanData <- stanData[,.(R=.N),by=.(timeNoVax,timeWithVax,ltbi)]
data = list(N=nrow(stanData),
            R=stanData$R,
            y=stanData$ltbi*stanData$R,
            timeNoVax=stanData$timeNoVax,
            timeWithVax=stanData$timeWithVax
            )

stan_code = "
data {
  int<lower=0> N;
  int R[N];
  int y[N];
  vector[N] timeNoVax;
  vector[N] timeWithVax;
}
parameters {
  real<lower=0,upper=0.5> thetaNoVax;
  real<lower=0,upper=0.5> thetaWithVax;
}
model {
  thetaWithVax ~ beta(0.05, 1);
  thetaNoVax ~ beta(0.05, 1);
  
  for (n in 1:N){
    y[n] ~ binomial(R[n], 1-((1-thetaWithVax)^timeWithVax[n])*((1-thetaNoVax)^timeNoVax[n]));
  }
}
"

pomp::bake("results/withVaxLTBI.RDS",{
  stan(model_code=stan_code,
              model_name="withVaxLTBI",
              data=data,
              iter=10000, chains=4, init=0, seed=4)
}) -> fit
summary(fit)$summary
```


## Detecting ATBI without vaccination

Lets try a hard model, detecting ATBI in unvaccinated people. ATBI is incident TB and detected in the year that it occurs.

We have a Poisson binomial distribution, where the probability of person $i$ not having ATBI is:

$$Pr(ATBI_i=0 | p_\text{atbi}, p_\text{ltbi}, T_i) = Pr(LTBI_i=0 | p_\text{atbi}, p_\text{ltbi}, T_i) + Pr(LTBI_i=1 \text{ without activation} | p_\text{atbi}, p_\text{ltbi}, T_i)$$
$$ = (1-p_\text{ltbi})^{T_i} + \Sigma_{t=1}^{t=T_i}Pr(\text{Incident LTBI infection at time }t,\text{ no activation})$$
$$ = (1-p_\text{ltbi})^{T_i} + \Sigma_{t=1}^{t=T_i} (1-p_\text{ltbi})^{t-1} \times p_\text{ltbi} \times (1-p_\text{atbi})^{T_i-t+1}$$

And the probability of person $i$ having LTBI is:

$$Pr(ATBI_i=1 | p_\text{atbi}, p_\text{ltbi}, T_i) = 1 - Pr(ATBI_i=0 | p_\text{atbi}, p_\text{ltbi}, T_i) $$

```{r}
cases <- list()
for(i in 1:14){
  cases[[i]] <- peopleNoVax[[i]][atbi==1]
}
cases <- rbindlist(cases)
controls <- peopleNoVax[[17]][atbi==0 & ageAtTB==0]

cases[,ageAtTB:=0]
controls[,ageAtTB:=0]

stanData <- rbind(controls[,.(R=.N),by=.(timeNoVax,timeWithVax,atbi)],cases[,.(R=.N),by=.(timeNoVax,timeWithVax,atbi)])
data = list(N=nrow(stanData),
            R=stanData$R,
            y=stanData$atbi*stanData$R,
            timeNoVax=stanData$timeNoVax)

stan_code = "
data {
  int<lower=0> N;
  int timeNoVax[N];
  int R[N];
  int y[N];
}
parameters {
  real<lower=0,upper=0.5> ltbiNoVax;
  real<lower=0,upper=0.5> atbiNoVax;
}
model {
real ps;
  ltbiNoVax ~ beta(0.05, 1);
  atbiNoVax ~ beta(0.05, 1);
  
  {
    int replicates;
    int personTime;
    real prob_ltbi_no_tb;

    for (n in 1:N){
      prob_ltbi_no_tb = (1-ltbiNoVax)^timeNoVax[n];
      personTime = timeNoVax[n];
      replicates = R[n];

      for(t in 1:personTime){
         prob_ltbi_no_tb = prob_ltbi_no_tb + ((1-ltbiNoVax)^(t-1))*ltbiNoVax*((1-atbiNoVax)^(personTime-t+1));
      }

      y[n] ~ binomial(R[n], 1-prob_ltbi_no_tb);
    }
  }
}

"

pomp::bake("results/noVaxATBI.RDS",{
  stan(model_code=stan_code,
              model_name="noVaxATBI",
              data=data,
              iter=10000, chains=4, init=0, seed=4)
}) -> fit
summary(fit)$summary
```

